\documentclass{amsart}

\usepackage{amssymb,amsmath}

\newcommand{\del}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ddel}[2]{\frac{\partial^2 #1}{\partial #2^2}}
\renewcommand{\d}[2]{\frac{d #1}{d #2}}

\linespread{1.5}

\title{Hyperbolic discounting model} \date{\today}


\begin{document}

\maketitle

\section{Elements of the model}
\label{sec:elements-model}

\subsection{Consumer}
\label{sec:consumer}

Imagine a representative consumer with quasi-linear preferences across
two goods: a polluting good, $x$, and a numeraire, $y$. The consumer's
instantaneous utility function is
\begin{equation}
  \label{eq:1} u = ax - \frac{b}{2}x^2 + y.
\end{equation} The consumer's budget constraint is
\begin{equation}
  \label{eq:2} p_x x + y = W,
\end{equation} hence, demand for $x$ is
\begin{equation}
  \label{eq:3} x = \frac{a}{b} - \frac{p}{b}
\end{equation} and inverse demand is
\begin{equation}
  \label{eq:4} p = a - bx.
\end{equation}

\subsection{Monopolist}
\label{sec:monopolist}

The manufacturer of good $x$ faces costs $cx$ and monopolises the
market. His instantaneous profit function is
\begin{equation}
  \label{eq:5} \pi = ax - bx^2 - cx.
\end{equation}

\subsection{Regulator}
\label{sec:regulator}

The regulator recognises the pollution damage, $\phi(k_t)$, caused by
the existence of the stock, $k_t$, of good $x$. The stock of the good
evolves according to the rule
\begin{equation}
  \label{eq:7} k_t = \theta k^{t-1} + x^{t-1} \qquad \theta \in [0,1].
\end{equation} Instantaneous welfare is given by
\begin{align}
  \label{eq:6} w_t &= \pi_t + CS_t - \phi(k_t) \\ &= \left[ ax - bx^2
- cx \right] + \left[ \left( a-p_t^* \right) \frac{x_t^*}{2} \right] -
\phi(k_t),
\end{align} where the superscripted asterisk denotes the equilibrium
value of the variable.

Let the pollution function be $\phi(k_t) = dk_t^2$. Now the
instantaneous welfare function can be written
\begin{equation}
  \label{eq:10} w_t = (a-c)x_t - \frac{b}{2}x_t^2 - dk_t^2.
\end{equation}

The regulator suffers from time inconsistency and is modelled as
having quasi-hyperbolic preferences with $\beta\delta$-discounting.

\section{Laissez Faire equilibrium}
\label{sec:laiss-faire-equil}

Suppose that the monopolist acted unregulated. He then solves a static
problem in each period:
\begin{align}
  \label{eq:8} MR &= MC \\ a-2bx &= c \\ x^* &= \frac{a-c}{2b} \\ p^*
  &= \frac{a+c}{2}.
\end{align} Thus
\begin{align}
  \label{eq:9} CS\left( x^*\right) &= \frac{\left( a-c \right)^2}{8b}
\\ \intertext{and} \pi\left( x^*\right) &= \frac{\left( a-c
\right)^2}{4b}.
\end{align}

In the dynamic, infinite horizon case these profit and CS values will
be summed to infinity with a discount rate of $\delta$. They will
simply be multiplied by $\frac{1}{1-\delta}$.Thus
\begin{align}
  \label{eq:9} CS\left( x^*\right) &= \frac{\left( a-c \right)^2}{8b}
\\ \intertext{and} \pi\left( x^*\right) &= \frac{\left( a-c
\right)^2}{4b}.
\end{align}


\section{Regulation with precommitment}
\label{sec:regul-with-prec}

Suppose that the regulator can directly choose $\{ x_\tau
\}_{\tau=t}^\infty$ at time $t$. The regulator must maximise
\begin{equation}
  \label{eq:11} W_t = \beta \sum_{i=1}^\infty \delta^i \left[
(a-c)x_{t+i} - \frac{b}{2}x_{t+i}^2 - dk_{t+i}^2 \right] + \left[
(a-c)x_t - \frac{b}{2}x_t^2 - dk_t^2 \right]
\end{equation} where $k_t = \theta k_{t-1} + x_{t-1}$ describes the
evolution of the state variable, $\delta$ is the discount rate and
$\beta$ is the quasi-hyperbolic modifier on the future discount rate.

Notating current instantaneous welfare in an abbreviaNotating current instantaneous welfare in an abbreviated fashion as
$w(x_t,k_t)$ allows the NPV of total future welfare to be written as
\begin{equation}
  \label{eq:30} W_0 = w(x_0,k_0) + \beta\delta w(x_1,\theta k_0+x_0) +
\beta\delta^2 w(x_2,x_1+\theta x_0+\theta^2 k_0) + \ldots
\end{equation} Maximising this requires taking first-order conditionss
with respect to $x_t \forall t\in [0,\infty)$. Derivatives are notated
as usual with the time superscript taken from the time subscript of
the state variable.
\begin{align}
  \label{eq:31} \del{W_0}{x_0} &= w_1^0 + \beta\delta w_2^1 +
\theta\beta\delta^2 w_2^2 + \ldots = 0 \\ \label{eq:32} \del{W_0}{x_1}
&=\beta\delta w_1^1 + \theta\beta\delta^2 w_2^2 + \theta\beta\delta^3
w_2^3 + \ldots = 0 \\ \label{eq:33} \del{W_0}{x_2} &=\beta\delta^2
w_1^2 + \theta\beta\delta^3 w_2^3 + \theta\beta\delta^4 w_2^4 + \ldots
= 0 \\ &\hspace{2cm}\vdots
\end{align} Compacting these conditions,
\eqref{eq:31}$-$\eqref{eq:32}$\times\theta$ gives
\begin{equation}
  \label{eq:34} w^0_1 + \beta\delta \left(w^1_2 - \theta w^1_1 \right)
= 0,
\end{equation} and $($\eqref{eq:33}$-$\eqref{eq:34}$\times\theta ) /
\beta\delta$ gives
\begin{equation}
  \label{eq:35} w^t_1 + \delta \left(w^t_2 - \theta w^{t+1}_1 \right)
= 0 \qquad \forall\; t \geq 2.
\end{equation}
ted fashion as
$w(x_t,k_t)$ allows the NPV of total future welfare to be written as
\begin{equation}
  \label{eq:12} W_0 = w(x_0,k_0) + \beta\delta w(x_1,\theta k_0+x_0) +
\beta\delta^2 w(x_2,x_1+\theta x_0+\theta^2 k_0) + \ldots
\end{equation} Maximising this requires taking first-order conditionss
with respect to $x_t \forall t\in [0,\infty)$. Derivatives are notated
as usual with the time superscript taken from the time subscript of
the state variable.
\begin{align}
  \label{eq:13} \del{W_0}{x_0} &= w_1^0 + \beta\delta w_2^1 +
\theta\beta\delta^2 w_2^2 + \ldots = 0 \\ \label{eq:14} \del{W_0}{x_1}
&=\beta\delta w_1^1 + \theta\beta\delta^2 w_2^2 + \theta\beta\delta^3
w_2^3 + \ldots = 0 \\ \label{eq:15} \del{W_0}{x_2} &=\beta\delta^2
w_1^2 + \theta\beta\delta^3 w_2^3 + \theta\beta\delta^4 w_2^4 + \ldots
= 0 \\ &\hspace{2cm}\vdots
\end{align} Compacting these conditions,
\eqref{eq:13}$-$\eqref{eq:14}$\times\theta$ gives
\begin{equation}
  \label{eq:16} w^0_1 + \beta\delta \left(w^1_2 - \theta w^1_1 \right)
= 0,
\end{equation} and $($\eqref{eq:15}$-$\eqref{eq:16}$\times\theta ) /
\beta\delta$ gives
\begin{equation}
  \label{eq:17} w^t_1 + \delta \left(w^t_2 - \theta w^{t+1}_1 \right)
= 0 \qquad \forall\; t \geq 2.
\end{equation}


\section{Regulation without precommitment}
\label{sec:regul-with-prec-2}

Now suppose that the regulator can still directly choose output but is
no longer able to precommit to future output decisions. The problem
must be formulated recursively in order to solve for a time consistent
output path.

The problem will first be solved with reduced form notation and then
functional forms can later be substituted in.

Let the MPE strategy of the regulator be $x_t = f(k_t)$. Then his
current period value function is
\begin{equation}
  \label{eq:18} U(k_t) = \max_{x_t} \left\{ w(x_t,k_t) + \beta\delta
V( \theta k_t + x_t) \right\}.
\end{equation} From period $t+1$ onward the value function is
$V(k_t)$.
\begin{equation}
  \label{eq:19} V(k_t) = w\left( f(k_t),k_t\right) + \delta V\left(
\theta k_t + f(k_t) \right).
\end{equation}

The current period's FOC is
\begin{align} w^t_1 + \beta\delta V^{t+1}_1 &= 0 \\ \label{eq:20}
\therefore V^{t+1}_1 &= -\frac{w^t_1}{\beta\delta}.
\end{align} Differentiating \eqref{eq:19} gives
\begin{equation}
  \label{eq:21} V^t_1 = w^t_1f^t_1 + w^t_2 + \delta V^{t+1}_1 (\theta
f^t_1)
\end{equation} and substituting in equation \eqref{eq:20} gives the
reduced form Euler-Lagrange equation:
\begin{equation}
  \label{eq:22} w^t_1 + \beta\delta( w^{t+1}_1 f^{t+1}_1 + w^{t+1}_2 )
- \delta (\theta + f^{t+1}_1)w^{t+1}_1 = 0.
\end{equation}

\section{Regulation with delegation}
\label{sec:regul-with-deleg}

The delegation game involves the regulator setting a tax rate for
pollution simultaneously with the monopolist's choice of output in
each period. Both the tax and the choice of output are feedback
strategies. We consider only a linear tax, however there are two
possible ways to levy it. It can be levied on either emissions or upon
the stock of pollution.

The case of a tax on emissions will be considered first and then the
problem with a pollution tax will be solved.

\subsection{The welfare function}
\label{sec:welfare-function}

The regulator's problem changes for two reasons: first, because he
gains revenue from taxation and, secondly, because we introduce a cost
to changing the tax rate. Economists are often criticised by
policymakers for excluding the costs of implementation when they
recommend taxes. Here, we explicitly include the costs of implementing
and modifying tax schemes in the regulator's welfare function.

Suppose that the tax is levied on emissions, the value of the tax
revenue to the regulator is $\gamma$, and the the cost of changing
policies is $\kappa (\tau_t-\tau_{t-1})^2$, where $\tau_t$ is the
period $t$ tax rate. Then the welfare function becomes
\begin{align}
  \label{eq:23} w_t &= (a-c)x_t - bx_t^2 + (\gamma -1)\tau_tx_t +
(a-p^*)\frac{x_t^*}{2} - dk_t^2 - \kappa (\tau_t-\tau_{t-1})^2 \\ &=
\omega(x_t,k_t) + (\gamma -1)\tau_tx_t - \kappa \rho(\tau_t,\tau_{t-1}).
\end{align} Note that, if $\gamma=1$ then the tax is a simple transfer
from the monopolist to the consumers and doesn't directly affect the
welfare function. Similarly, if $\kappa=0$ then there are no costs of
policy adjustment.

\subsection{The profit function}
\label{sec:profit-function}

With taxation the monopolist's instantaneous profit becomes
\begin{equation}
  \label{eq:24} \pi_t = (a-c-\tau_t)x_t - bx_t^2.
\end{equation}

\subsection{The game}
\label{sec:game}

Let the MPE strategy of the monopolist be $x_t = h(\tau_{t-1},k_t)$
and the MPE strategy of the regulator be $\tau_t = g(\tau_{t-1},k_t)$.

\subsubsection{Regulator}
\label{sec:regulator-1}

Current period value function:
\begin{multline}
  \label{eq:25} U(\tau_{t-1},k_t) = \max_{\tau_t} \bigg\{ \omega\big(
h(\tau_{t-1},k_t),k_t\big) + (\gamma -1)\tau_tx_t - \kappa \rho(\tau_t,\tau_{t-1}) \\ + \beta\delta V \big(
\tau_t, \theta k_t + h(\tau_{t-1},k_t)\big) \bigg\}.
\end{multline} Continuation value function:
\begin{multline}
  \label{eq:26} V(\tau_{t-1},k_t) = \omega\left(
h(\tau_{t-1},k_t),k_t\right)  + (\gamma -1)g(\tau_{t-1},k_t)x_t
- \kappa \rho(\tau_t,\tau_{t-1}) \\+
\delta V \left( g(\tau_{t-1},k_t), \theta k_t +
h(\tau_{t-1},k_t)\right).
\end{multline} FOC:
\begin{align}
  (\gamma-1)x_t + \kappa\rho^t_1 + \beta\delta V^{t+1}_1 &= 0 \\ \label{eq:27}
  V^{t+1}_1 &= -\frac{(\gamma-1)x_t + \kappa\rho^t_1}{\beta\delta}.
\end{align} Envelope conditions from \eqref{eq:26}:
\begin{align}
  \label{eq:28} V^t_1 &= \omega^t_1h^t_1 + (\gamma-1)x_tg^t_1 - \kappa\rho^t_2 + \delta
V^{t+1}_1g^t_1 + \delta V^{t+1}_2h^t_1 \\ \label{eq:29} V^t_2 &= \omega^t_2 +
\omega^t_1h^t_2 + (\gamma-1)x_tg^t_2 + \delta V^{t+1}_1g^t_2 + \delta
V^{t+1}_2 (\theta + h^t_2).
\end{align} Solving for the Euler-Lagrange equation: \eqref{eq:27}
$\rightarrow$ \eqref{eq:28} gives
\begin{gather}
  \label{eq:30} \omega^t_1h^t_1 + (\gamma-1)x_tg^t_1 -\kappa\rho^t_2 +
  \delta \left(
    V^{t+1}_2h^t_1 - g^t_1\frac{(\gamma-1)x_t + \kappa\rho^t_1}{\beta\delta} \right) + \frac{(\gamma-1)x_{t-1} + \kappa\rho^{t-1}_1}{\beta\delta} = 0 \\
  \label{eq:31} \therefore V^{t+1}_2 = \frac{g^t_1 ( \kappa \rho^t_1
    + (\gamma-1)x_t)}{\beta\delta h^t_1} - \frac{\kappa\rho^{t-1}_1 +
    (\gamma-1)x_{t-1}}{\beta\delta h^t_1} + \frac{\kappa\rho^t_2 -
    (\gamma-1)x_tg^t_1}{\delta h^t_1} - \frac{\omega^t_1}{\delta}.
\end{gather} Now \eqref{eq:27},~\eqref{eq:31}
$\rightarrow$~\eqref{eq:29} gives the result
\begin{multline}
  \label{eq:32}
  h_1^t \kappa \rho_2^{t-1} \beta \delta- h_1^t \omega_1^{t-1}
  h_1^{t-1} \beta \delta- h_1^t g_1^{t-1} \delta x^{t-1}+ h_1^t
  x^{t-2}\\-\delta h_1^{t-1} h_2^t \kappa \rho_1^{t-1}- \delta ^2
  h_1^{t-1} \kappa \rho_2^t \beta \theta-{ \delta}^2 h_1^{t-1} h_2^t
  \kappa \rho_2^t \beta+ \delta ^{ 2} h_1^{t-1} h_2^t g_1^t x^t\\ +
  h_1^t g_1^{t-1} x^{t-1} \beta \delta- \delta ^2 h_1^{t-1} h_2^t
  g_1^t x^t \beta+ \delta ^2 h_1^{t-1} g_1^t \kappa \rho_1^t \theta+
  \delta ^2 h_1^{t-1} h_2^t g_1^t \kappa \rho_1^t + \delta ^2
  h_1^{t-1} g_1^t x^t \theta\\+ \delta ^2 h_1^{t-1} g_1^t x^t \beta
  \gamma h_2^t- \delta ^2 h_1^{t-1} g_1^t x^t \gamma h_2^t- h_1^t
  g_1^{t-1} x^{t-1} \beta \delta \gamma+ h_1^t \kappa \rho_1^{t-2}-
  h_1^t g_1^{t-1} \delta \kappa \rho_1^{t-1}\\+ g_2^t h_1^t {
    \delta}^2 h_1^{t-1} x^t \gamma- x^t g_2^t \beta h_1^t \delta ^2
  h_1^{t-1} \gamma+\delta h_1^{t-1} x^{t-1} \gamma \theta+\delta
  h_1^{t-1} x^{t-1} \gamma h_2^t- \delta ^2 h_1^{t-1} g_1^t x^t \gamma
  \theta\\+ \delta ^2 h_1^{t-1} g_1^t x^t \beta \gamma \theta+ h_1^t
  g_1^{t-1} \delta x^{t-1} \gamma- g_2^t h_1^t \delta ^2 h_1^{t-1}
  \kappa \rho_1^t- \omega_2^t \beta h_1^t \delta ^2 h_1^{t-1}+ x^t
  g_2^t \beta h_1^t \delta ^2 h_1^{t-1}\\- \delta h_1^{t-1} \kappa
  \rho_1^{t-1} \theta- g_2^t h_1^t { \delta}^2 h_1^{t-1} x^t + \delta
  ^2 h_1^{t-1} \omega_1^t h_1^t \beta \theta- \delta ^2 h_1^{t-1}
  g_1^t x^t \beta \theta-\delta h_1^{t-1} h_2^t x^{t-1}\\-\delta
  h_1^{t-1} x^{t-1} \theta- h_1^t x^{t-2} \gamma{\beta h_1^t {\delta
    }^2 h_1^{t-1}} = 0
\end{multline}
Note that, if $\gamma=1, \kappa=0$ then this simplifies to
\begin{equation}
\label{eq:48}
something
\end{equation}


\subsubsection{Monopolist}
\label{sec:monopolist-1}

Since the monopolist discounts exponentially, he has a stationary
value function:
\begin{equation}
  \label{eq:33}
  \Pi(\tau_{t-1},k_t) = \max_{x_t} \left\{
    \pi(x_t,g\left(\tau_{t-1},k_t)\right) + \delta \Pi \left(
      g(\tau_{t-1},k_t), \theta k_t + x_t \right) \right\}.
\end{equation}
FOC:
\begin{align}
  \pi^t_1 + \delta \Pi^{t+1}_2 &= 0 \\  \label{eq:34}
  \therefore \Pi^{t+1}_2 &= -\frac{\pi^t_1}\delta.
\end{align}
Envelope conditions:
\begin{align}
  \label{eq:35}
  \Pi^t_1 &= \pi^t_2g^t_1 + \delta \Pi^{t+1}_1g^t_1 \\\label{eq:36}
  \Pi^t_2 &= \pi^t_2g^t_2 + \delta \Pi^{t+1}_1g^t_2 + \theta\delta\Pi^{t+1}_2.
\end{align}
Solving for the Euler equation:
\eqref{eq:34}$\to$\eqref{eq:36} gives
\begin{align}
  \label{eq:37}
  -\frac{\pi^{t-1}_1}\delta &= \pi^t_2g^t_2 - \theta\pi^t_1 + \delta
  \Pi^{t+1}_1g^t_2 \\\label{eq:38}
  \Pi^{t+1}_1 &= \frac{\theta\pi^t_1 - \pi^t_2g^t_2 -
    \frac{1}\delta\pi^{t-1}_1}{\delta g^t_2}.
\end{align}
Now, (\ref{eq:38})$\rightarrow$(\ref{eq:35}) yields the Euler-Lagrange
equation:
\begin{equation}
  \label{eq:39}
  \pi^t_2g^t_1 + \frac{g^t_1}{g^t_2} \left( \theta \pi^t_1 -
    \pi^t_2g^t_2 - \frac{1}\delta\pi^{t-1}_1 \right) -
  \frac{\theta\pi^{t-1}_1 - \pi^{t-1}_2g^{t-1}_2 -
    \frac{1}\delta\pi^{t-2}_1}{\delta g^{t-1}_2} = 0
\end{equation}


\subsection{Taxation of the pollution stock}
\label{sec:taxat-pollut-stock}

\subsubsection{Regulator}
\label{sec:regulator-2}

The instantaneous welfare function becomes
\begin{align}
  \label{eq:47}
  w_t &= (a-c)x_t - bx_t^2 + (\gamma -1)\tau_tk_t +
  (a-p^*)\frac{x_t^*}{2} - dk_t - \kappa (\tau_t-\tau_{t-1})^2 \\ &=
  w(x_t,\tau_t,\tau_{t-1},k_t).
\end{align}
Since the reduced form of the welfare function is the same, the
reduced form Euler equation wil be the same as equation (\ref{eq:32}).


\subsubsection{Monopolist}
\label{sec:monopolist-2}

The instantaneous profit function becomes
\begin{align}
  \label{eq:40}
  \pi_t &= (a-c)x_t - bx_t^2 - \tau_tk_t \\
  &= \pi(x_t,\tau_t,k_t).
\end{align}
The Bellman function is then
\begin{equation}
  \label{eq:41}
  \Pi(\tau_{t-1},k_t) = \max_{x_t} \left\{ \pi\left(
      x_t,g(\tau_{t-1},k_t),k_t\right) + \delta \Pi \left(
      g(\tau_{t-1},k_t), \theta k_t + x_t \right) \right\}.
\end{equation}
FOC:
\begin{equation}
  \label{eq:42}
  \Pi^{t+1}_2 = -\frac{\pi^t_1}\delta.
\end{equation}
Envelope conditions:
\begin{align}
  \label{eq:43}
  \Pi^t_1 &= \pi^t_2g^t_1 + \delta \Pi^{t+1}_1g^t_1 \\ \label{eq:45}
  \Pi^t_2 &= \pi^t_2g^t_2 + \pi^t_3 + \delta \Pi^{t+1}_1g^t_2 + \delta
  \theta \Pi^{t+1}_2.
\end{align}
Solving for the Euler equation:
(\ref{eq:42})$\rightarrow$~(\ref{eq:45}) rearranges to
\begin{equation}
  \label{eq:44}
  \Pi^{t+1}_1 = \frac{\theta \pi^t_1 - \pi^t_3 - \pi^t_2g^t_2 -
    \frac{1}\delta\pi^{t-1}_1}{\delta g^t_2}
\end{equation}
and then substituting (\ref{eq:44}) into (\ref{eq:43}) gives the
Euler-Lagrange equation:
\begin{multline}
  \label{eq:46}
  \pi^t_2g^t_1 + \frac{g^t_1}{g^t_2} \left( \theta \pi^t_1 -
    \pi^t_2g^t_2 - \pi^t_3 - \frac{1}\delta\pi^{t-1}_1 \right) \\ -
  \frac{\theta\pi^{t-1}_1 - \pi^{t-1}_2g^{t-1}_2 - \pi^t_3 -
    \frac{1}\delta\pi^{t-2}_1}{\delta g^{t-1}_2} = 0
\end{multline}







\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
