\documentclass[12pt, a4paper, oneside, openright]{book}

\usepackage{vuwthesis} % sets up some local things, mostly the front page
\usepackage{palatino} % sets palatino as the default font
%\usepackage{url} % for typesetting urls
\usepackage{amssymb, amsmath, amsthm}
\usepackage[round]{natbib}
\usepackage{graphicx}
\usepackage{rotating}
\usepackage{subfig}
\usepackage{multirow}

\newcommand{\del}[2]{\frac{\partial #1}{\partial #2}}
\newcommand{\ddel}[2]{\frac{\partial^2 #1}{\partial #2^2}}
\renewcommand{\d}[2]{\frac{d #1}{d #2}}

\newcommand{\vect}[1]{\mathbf{#1}}

\theoremstyle{plain}
\newtheorem{thm}{Theorem}
\newtheorem{prop}{Proposition}

\theoremstyle{definition}
\newtheorem{defn}{Definition}
\newtheorem{ass}{Assumption}

\linespread{1.5}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\frontmatter

\title{Environmental regulation of firms that experience dynamic
  inconsistency}
\author{James Zuccollo}

\subject{Economics}

\abstract{The recent push for environmental regulation has invigorated
  the discussion of mechanism design and optimal taxation
  policy. Recent decades have also seen growing interest in
  behavioural economics and empirically based theory. In this thesis
  we take a step towards combining the two by asking how a regulator
  may correct an externality in situations where they have a time
  consistency problem.

  Time inconsistency is one of the notable developments of behavioural
  economics. It posits that an agent's decisions do not remain
  consistent over time, which causes a utility loss if the agent
  cannot commit themselves to a particular course of action and stick
  to it. The solution to inconsistency problems is to precommit to a
  course of action and prevent future deviations from it. However,
  finding a mechanism to enable such precommitment is often
  problematic.

  A regulator who maximises welfare can have a time consistency
  problem because welfare will depend on the decisions of firm and
  households who may themselves be inconsistent. That inconsistency
  then propagates to the regulator's decision and reduces the level of
  welfare that the regulator can reach. Alternatively, the regulator's
  time consistency problem can be caused by non-stationarity in their
  time preferences.  To reach the first-best outcome the regulator
  must not only eliminate the environmental externality: they must
  also overcome their own time inconsistency problem.

  This thesis draws from the literature on strategic delegation to
  construct a taxation game in which the regulator can achieve the
  first best taxation regime without the need for external
  precommitment devices. We study a dynamic game where the regulator
  chooses a tax rate and the regulated monopolist chooses their
  price. We show that the Markov-perfect equilibrium price path of
  this game will replicate the first best plan. Our results holds for
  time inconsistency caused by both jump states and quasi-hyperbolic
  discounting.}

\otherdegree{Master of Commerce and Administration}

\maketitle

\include{acknowledge}

\tableofcontents

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\mainmatter

\include{introduction}

\include{theory}

\include{jump_problem}

\include{jump_proposition}

\include{jump_computation}

\include{hyperbolic}

\include{conclusion}

\appendix

\include{assump_app}

\include{durable_app}

\include{jump_comp_appendix}

\include{hyp_appendix}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\bibliographystyle{ieeetr} %originally used by vuwthesis style
\bibliographystyle{plainnat}
\bibliography{Thesis}


\end{document}
